
            output:
            html_document: default
            pdf_document: default
            
============================================================================

============================================================================
                  
                  Title: "Harmful Weather Events"
                  Author: "Christopher"
                  Date: "July 21, 2019"
                  Output: html_document
                  
============================================================================
                          
                          SYNOPSIS
                          
============================================================================
Across the United States, excessive heat,tornadoes and flash floods are most harmful with respect to population health.

Across the United States, tornadoes,thunderstorm, winds and flash floods have the greatest  economic consequences.

Our raw data is taken from 
[National Weather Service Instruction 10-1065][1].
The events in the database start in the year 1950 and end in November 2011.
Injuries,Fatalities, Property Damage and Crop Damage(in Dollars)are calculated during that times.

[1]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

```{r setup, include=FALSE}


##Data processing

library(ggplot2)
library(dplyr)
library(utils)
library(plyr)
library(markdown)
library(knitr)
devtools::install_github("rstudio/rsconnect")
```
## Downloads and reads the dataset into R
```{r cache = TRUE}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url,"StormData.csv.bz2")
data <- read.csv("C:/Users/Hlangano/Downloads/repdata_data_StormData (4).csv")
head(data)
```
## Subsetting the storm data in memory
```{r}
Subdata <- c("EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGExP")
```
## Convert B,M,K,H units to calculate Property Damage


 
 ## Convert B,M,K,H units to calculate Crop damage
 
 ```{r}
 Subdata$CropDamage <- 0
 Subdata[Subdata$CROPDMGEXP == "B"]$CropDamage <- Subdata[Subdata$CROPDMGEXP == "B"]$CROPDMG*10^9
 Subdata[Subdata$CROPDMGEXP == "M"]$CropDamage <- Subdata[Subdata$CROPDMGEXP == "M"]$CROPDMG*10^6
 Subdata[Subdata$CROPDMGEXP == "K"]$CropDamage <- Subdata[Subdata$CROPDMGEXP == "K"]$CROPDMG*10^3
 Subdata[Subdata$CROPDMGEXP == "H"]$CropDamage <- Subdata[Subdata$CROPDMGEXP == "H"]$CROPDMG*10^2
```
## What causes most injuries
```{r}
injuries <- aggregate(data$INJURIES,by = list(EVTYPE = data$EVTYPE),sum)
 injuries <- injuries[order(injuries$x,decreasing = TRUE), ]
 head(injuries)
 
 ```
 ## Plot for most injuries caused
 
 ```{r}
  ggplot(injuries[1:5,], aes(EVTYPE, y= x)) + geom_bar(stat ="identity") + xlab("Events Type")+ ylab("Number of Injuries")+ggtitle("Injuries by Event Type")
```
  ## The 5 most fatalities events
  
```{r}
   fatalities <- aggregate(data$FATALITIES,by = list(EVTYPE = data$EVTYPE),sum)
 fatalities <- fatalities[order(fatalities$x,decreasing = TRUE), ]
head(fatalities)

```

 ## Plot for fatalities
```{r,echo = TRUE, cache = TRUE}

 
 ggplot(fatalities[1:5,], aes(EVTYPE, y= x)) + geom_bar(stat ="identity") + xlab("Events Type")+ ylab("Number of Fatalities")+ggtitle("Injuries by Event Type")

```
 ## We combined the exponents with the value
 
```{r,echo = TRUE, cache=TRUE}
Subdata$PROPDMGEXP <- 10^(as.numeric(Subdata$PROPDMGEXP))
Subdata$CROPDMGEXP <- 10^(as.numeric(Subdata$CROPDMGEXP))
```

 ## The top 5 events which the highest total economic damages
```{r,echo=TRUE, echo= TRUE}
Subdata$CROPEXP[Subdata$CROPDMGEXP ==""] <- 1

 # Assigning "0" to invalid exponent data
Subdata$CROPEXP[Subdata$CROPDMGEXP == "?"] <- 0
```
## Calculating the property damage value
```{r, echo=TRUE, cache= TRUE }
Subdata$CROPDMGVAL <- Subdata$CROPDMG*Subdata$CROPEXP
 # Assigning "0" to invalid exponet data
Subdata$PROPEXP[Subdata$PROPDMGEXP ==""] <- 1
Subdata$PROPEXP[Subdata$PRPODMGEXP =="+"] <- 0
Subdata$PROPEXP[Subdata$PRPODMGEXP =="+"] <- 0
 Subdata$PROPEXP[Subdata$PRPODMGEXP =="?"] <- 0

Subdata$PROPEXP[Subdata$PRPODMGEXP =="+"] <- 0
 Subdata$PROPEXP[Subdata$PRPODMGEXP =="?"] <- 0
```
## Results
### Damage to human life
Tornadoes caused most injuries and fatalities

### Damage to Property and Crop

Floods are responsible for the most economic damage.
 
 
 
 
